---
title: "Importing Data in R: Part 2 - DataCamp Course Notes"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These notes and exercises are based on the DataCamp course Importing Data in R: Part 2. All notes and completed exercises are by Amy Gill. Exercises were originally designed by DataCamp and data sets are downloaded from the DataCamp servers.

## HTTP and HTTPS Compatibility: Importing Data from the Web

The built-in *read.table* (*read.csv*, *read.tsv*) and *readr* package *read_delim* functions (*read_csv*, *read_tsv*) are able to make http and https connections to import flat files from the web:

```{r}
# Load the readr package
library(readr)

# Import the csv file: pools
url_csv <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/swimming_pools.csv"
pools <- read_csv(url_csv)

# Import the txt file: potatoes
url_delim <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/potatoes.txt"
potatoes <- read_tsv(url_delim)

# Print pools and potatoes
pools
potatoes
```

## Using URLs to download HTTPS data sets including .RData

Some packages are not able to directly interface with HTTP/HTTPS and cannot directly download data from a URL. For example, the *readxl* package does not currently support direct data download. In this case, you will have to download data onto your local system and then import from the working directory.

Some file formats also do not have common package support for direct download. For example, .RData is a very efficient format for storing R Data that can also be imported via *download.file* and then loaded into the R session with *load*.

```{r}
# https URL to the wine RData file.
url_rdata <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/wine.RData"

# Download the wine file to your working directory
download.file(url_rdata, "wine_local.RData")

# Load the wine data into your workspace using load()
load("wine_local.RData")

# Print out the summary of the wine data
summary(wine)
```

## The httr package and the GET request

*GET* is an explicit HTTP GET call within R. This allows access to the raw html in addition to processed text and formatted objects such as JSON objects and arrays (see next section).

```{r}
# load httr
library(httr)

# Get the url
url <- "http://www.omdbapi.com/?apikey=ff21610b&t=Annie+Hall&y=&plot=short&r=json"
resp <- GET(url)

# Print resp
resp

# Print content of resp as text
content(resp, as = "text")

# Print content of resp
content(resp)
```

## The JSON Format, APIs, and the jsonlite Package

JSON format:simple, concise, well-structured; human-readable; easy to parse and generate for computers; great for communicating an information summary

API = application programming interface: set of routines and protocols for building software regulating how different components interact

A web API will generally be an interface to get/add data to server and manage HTTP "verbs" (GET and others)

For example, Twitter has an API:

https://dev.twitter.com/rest/public

This can allow many remote interactions with Twitter, including:
-get tweets
-place comments on tweets
-many applications including researching effects of tweets
  
The *jsonlite* package is designed for conveniently working with JSON formats and improves on earlier R packages by providing consistent, robust functions that support all use cases.

JSONs can be built in R with curly braces as {"key":"value"} pairs and extracted into lists with *fromJSON*:

```{r}
# Load the jsonlite package
library(jsonlite)

# wine_json is a JSON
wine_json <- '{"name":"Chateau Migraine", "year":1997, "alcohol_pct":12.4, "color":"red", "awarded":false}'

# Convert wine_json into a list: wine
wine <- fromJSON(wine_json)

# Print structure of wine
str(wine)
```

JSONs can also be extracted directly from URLs:

```{r}

# Definition of quandl_url
quandl_url <- "https://www.quandl.com/api/v3/datasets/WIKI/FB/data.json?auth_token=i83asDsiWUUyfoypkgMz"

# Import Quandl data: quandl_data
quandl_data <- fromJSON(quandl_url)

# Print structure of quandl_data
str(quandl_data)
```

Data extracted into R from JSON format objects can be easily queried and compared:

```{r}

# Definition of the URLs
url_sw4 <- "http://www.omdbapi.com/?apikey=ff21610b&i=tt0076759&r=json"
url_sw3 <- "http://www.omdbapi.com/?apikey=ff21610b&i=tt0121766&r=json"

# Import two URLs with fromJSON(): sw4 and sw3
sw4 <- fromJSON(url_sw4)
sw3 <- fromJSON(url_sw3)

# Print out the Title element of both lists
sw4$Title
sw3$Title


# Is the release year of sw4 later than sw3?
sw4$Year > sw3$Year
```
```{r}

# Challenge 1
json1 <- '[1, 2, 3, 4, 5, 6]'
fromJSON(json1)

# Challenge 2
json2 <- '{"a": [1, 2, 3], "b":[4, 5, 6]}'
fromJSON(json2)
```

```{r}
# jsonlite is already loaded

# Challenge 1
json1 <- '[[1, 2], [3, 4]]'
fromJSON(json1)

# Challenge 2
json2 <- '[{"a": 1, "b": 2}, {"a": 3, "b": 4}, {"a":5, "b":6}]'
fromJSON(json2)
```

You can also make your own JSON objects from other R objects using *toJSON*.

By default, the display will be in mini format, which is compact but difficult for the human eye to read. This can be changed with the argument *pretty = TRUE*.

Existing JSON objects can be switched between pretty and mini display format using *prettify* and *minify*.

```{r}
# jsonlite is already loaded

# URL pointing to the .csv file
url_csv <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/water.csv"

# Import the .csv file located at url_csv
water <- read.csv(url_csv, stringsAsFactors = FALSE)

# Convert the data file according to the requirements
water_json <- toJSON(water)

# Print out water_json
water_json
```
```{r}
# jsonlite is already loaded

# Convert mtcars to a pretty JSON: pretty_json
pretty_json <- toJSON(mtcars, pretty = TRUE)

# Print pretty_json
pretty_json

# Minify pretty_json: mini_json
mini_json <- minify(pretty_json)

# Print mini_json
mini_json
```

## Importing Data from Statistical Software (SAS, STATA, SPSS) - The haven Package

Several statistical software suites are used in research and business, and these data formats can also be imported in R. Three common statistical software offerings include SAS, STATA, and SPSS.

SAS - Statistical Analysis Software - business analytics, biostatistics, medical sciences - .sas7bdat, .sas7bcat

STATA - Statistics and Data - economics - .dta

SPSS - Statistical Packages for Social Sciences - social sciences - .sav, .por

*haven* is a package for importing SAS, STATA, and SPSS files.

In *haven*, SAS files are read with *read_sas*:

```{r}
# Load the haven package
library(haven)

# Import sales.sas7bdat: sales
sales_url <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/sales.sas7bdat"
sales <- read_sas(sales_url)

# Display the structure of sales
str(sales)
```

In *haven*, STATA files are read with *read_stata* or *read_dta*.

Note that some variables from STATA may be of the "labelled" class. If so, they will initially appear as integers with layered label data. The labels can be extracted with *as_factor*.

```{r}
# haven is already loaded

# Import the data from the URL: sugar
sugar_url <- "http://assets.datacamp.com/production/course_1478/datasets/trade.dta"
sugar <- read_dta(sugar_url)

# display sugar Date column
head(sugar$Date)

# Convert values in Date column to dates
sugar$Date <- as.Date(as_factor(sugar$Date))

# display converted sugar Date column
head(sugar$Date)
```

In *haven*, SPSS files can be read using *read_sav* or *read_por* based on their suffix.

SPSS files can also have "labelled" class columns that may need to be converted for R work with *as_factor*.

```{r}
# haven is already loaded

# Import SPSS data from the URL: work
work <- read_sav("http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/employee.sav")

# Display summary of work$GENDER
summary(work$GENDER)


# Convert work$GENDER to a factor
work$GENDER <- as_factor(work$GENDER)


# Display summary of work$GENDER again
summary(work$GENDER)

```

## Alternative Importing of Statistical Package Software Data - The foreign Package

*foreign* is a package written by the core R team that can import data from a variety of statistical programs, not just SAS/STATA/SPSS. However, its usage is less consistent and it has some compatibility issues (as of course writing in late 2016). This is a brief introduction to *foreign* and some of its arguments specifically for SAS/STATA/SPSS import.

To load STATA data, use *read.dta*:

```{r}
# Load the foreign package
library(foreign)

# Import florida.dta and name the resulting data frame florida
florida_url <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/florida.dta"
florida <- read.dta(florida_url)

# Check tail() of florida
tail(florida)
```

The key default for STATA is *convert.factors = TRUE*, which avoids the need for factor conversion after import.

```{r}
# foreign is already loaded

# Specify the URL for educational equality data
path <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/edequality.dta"

# Create and print structure of edu_equal_1
edu_equal_1 <- read.dta(path)
str(edu_equal_1)

# Create and print structure of edu_equal_2
edu_equal_2 <- read.dta(path, convert.factors = FALSE)
str(edu_equal_2)


# Create and print structure of edu_equal_3
edu_equal_3 <- read.dta(path, convert.underscore = TRUE)
str(edu_equal_3)

```

To import SPSS, use *read.spss*. By default, the data is read as a list rather than a data frame. This can be changed with the option *to.data.frame = TRUE*. To change whether labelled are changed to factors, alter *use.value.labels* from TRUE (default) to FALSE.

```{r}
# foreign is already loaded

# Import international.sav as demo_1
demo_1 <- read.spss("https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/international.sav", to.data.frame = TRUE)

# Print out the head of demo_1
head(demo_1)

# Import international.sav as demo_2
demo_2 <- read.spss("https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/international.sav", to.data.frame = TRUE, use.value.labels = FALSE)

# Print out the head of demo_2
head(demo_2)
```

More information can be found in the *foreign* documentation. In general, though, Hadley Wickham's *haven* is more intuitive and simpler if working with SAS/STATA/SPSS.

## Enjoy seamlessly importing data in R!