---
title: "Importing Data in R"
output: html_notebook
---

## Importing flat files - .csv, .txt, including odd delimiters

read.table("filename.txt", header = FALSE, sep = "/", stringsAsFactors = TRUE)
*header = TRUE* indicates first row lists variable names - default FALSE
plenty of other arguments like col.names, colClasses


For common delimited file types, wrapper functions with common defaults are supplied:

read.csv() - comma delimited
read.delim() - tab delimited


COMMA DELIMITED:
read.csv("filename.csv")
by default for read.csv *header = TRUE* and *sep = ","* 
also by default *stringsAsFactors = TRUE*, can set to FALSE to have them be character strings instead -

TAB DELIMITED:
read.delim("filename.txt")
by default *header = TRUE, sep = "\t", stringsAsFactors = TRUE*

There are also read.csv2() and read.delim2() to account for regional differences in representing numerals with commas/decimals - see documentation, and if you are not getting a clean import try the other function variant. 



```{r}

hotdogs <- read.delim("hotdogs.txt",header = FALSE, col.names = c("type", "calories", "sodium"))

lily <- hotdogs[which.min(hotdogs$calories), ]
lily

tom <- hotdogs[which.max(hotdogs$sodium), ]
tom

```

You can also specify the column classes of the resulting data frame with the *colClasses* argument:

read.delim("filename.txt", colClasses = c("character", "numeric", "logical"))

This is useful if some of your columns should be factors and others should be characters and can be used instead of stringsAsFactors.

If a column is set to "NULL" in colClasses, it will be skipped and not loaded into the data frame.


## The readr package

Fast, easy to use, consistent - compared to built-in utils, which is verbose and can be slow

Imports as tibbles

```{r}
library(readr)
```

In readr:
read_delim("filename.txt", delim = "/")
read_csv("filename.csv")
read_tsv("filename.txt")
By default, strings are not converted to factors, so there is no need for the stringsAsFactors argument.

For read_delim:
*col_names = TRUE* by default
*col_types* automatically inferred by first row, or can be manually specified using the following shorthand:
c=character d=double i=integer l=logical
so *col_types = "ccddil"* makes 6 columns, chr chr dbl dbl int lgl

Other useful functionality for readr read_delim:
skip = skip n rows at start before reading - remember add col_names if skipping header!
n_max = number of rows to read

## Reading Excel Format Files - The readxl Package

readxl has two main functions:
excel_sheets(): lists different sheets
read_excel(): import excel doc from .xls or .xlsx

```{r}
library(readxl)
```

*excel_sheets("cities.xlsx")* returns 
[1] "year_1990" "year_2000"

By default, read_excel() imports the first sheet unless told otherwise:
*read_excel("cities.xlsx")* imports sheet "year_1990"
*read_excel("cities.xlsx", sheet = 2)* imports sheet "year_2000"
*read_excel("cities.xlsx", sheet = "year_2000")* imports sheet "year_2000"

You can chain the two readxl functions with lapply to quickly import all sheets:
```{r}
## THIS FILE IS NOT IN THE WORKING DIRECTORY - SNIPPET ONLY
lapply(excel_sheets("cities.xlsx"), read_excel, path="cities.xlsx")
```
col_names can be TRUE (default) - first row of table, FALSE - R assigns placeholders, character vector with names

col_types can be "NULL" (R will guess), "text", "numeric", "date", or "blank" (R will not import that column)

ALSO AVAILABLE: gdata package - readxl is still under development, so if you are not comfortable using something before it's finalized-ish, learn gdata. The gdata *read.xls()* function is an intuitive extension of read.table, BUT it only works for .xls and not .xlsx without a patch.
```{r}
## NOTE THIS FILE IS NOT IN MY WORKING DIRECTORY - JUST A CODE SNIPPET


library(gdata)

# Column names for urban_pop
columns <- c("country", paste0("year_", 1967:1974))

# Finish the read.xls call
urban_pop <- read.xls("urbanpop.xls", sheet = 2,
                      skip = 50, header = FALSE, stringsAsFactors = FALSE,
                      col.names = columns)

# Print first 10 observation of urban_pop
head(urban_pop, 10)

```

## The XLConnect Package for Reproducible Excel Work

creates workbook objects - bridge between Excel and R
- Load with *loadWorkbook*
- List sheets with *getSheets*
- Read worksheet with *readWorksheet*
```{r}
## NOTE THIS FILE IS NOT IN MY WORKING DIRECTORY - JUST A CODE SNIPPET

library(XLConnect)

book <- loadWorkbook("cities.xlsx")
getSheets(book)

readWorksheet(book, sheet = 2)
readWorksheet(book, sheet = "year_2000")
```

The startRow, endRow, startCol, endCol args can define selective inputs:
```{r}
readWorksheet(book, sheet = 2, startRow = 3, endRow = 4, startCol = 2, header = FALSE)
```

- Add sheets with *createSheet*, *writeWorksheet*
- Save changes with *saveWorkbook*
- Rename sheets with *renameSheet*
- Remove sheets with *removeSheet*
- Plenty more to style cells, arrange, and more...
```{r}
createSheet(book, name = "year_2010")
pop_2010 <- ## TRUNCATED
writeWorksheet(book, pop_2010, sheet = "year_2010")
saveWorkbook(book, file = "cities2.xlsx")
renameSheet(book, year_2010, Y2010)
removeSheet(book, Y2010)
```

