---
title: "MBTA Ridership Data Analysis"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Initial cleaning and analysis based on the DataCamp course "Importing and Cleaning Data in R: Case Studies" and extended by Amy Gill.
Data was downloaded from the DataCamp course repository.

```{r}
# Load readxl
library(readxl)

# Import mbta.xlsx and skip first row: mbta
mbta <- read_excel("data/mbta.xlsx", skip = 1)
```
```{r}
str(mbta)
```

```{r}
head(mbta, n = 20)
```

First, remove unnecessary rows and columns. Some information in this table is not relevant to monthly MBTA ridership changes (quarterly, pct chg/yr, TOTAL, row numbers).

```{r}
# Remove rows 1, 7, and 11 of mbta: mbta2
mbta2 <- mbta[c(-1, -7, -11), ]

# Remove the first column of mbta2: mbta3
mbta3 <- mbta2[, -1]
```

Next, gather columns except mode into a month column with value thou_riders (thousand riders).

```{r}
## mbta3 is pre-loaded

# Load tidyr
library(tidyr)

# Gather columns of mbta3: mbta4
mbta4 <- gather(mbta3, month, thou_riders, -mode)

# View the head of mbta4
head(mbta4)
```

Coerce riders to numeric, then split different transit options into separate columns.

```{r}
# Coerce thou_riders to numeric
mbta4$thou_riders <- as.numeric(mbta4$thou_riders)

# Spread the contents of mbta4: mbta5
mbta5 <- spread(mbta4, mode, thou_riders)

head(mbta5)
```

Split month column into month and year.

```{r}
# Separate month column into month and year: mbta6
mbta6 <- separate(mbta5, month, c("year", "month"))

head(mbta6)
```

Check distributions of variables with *summary*.

```{r}
summary(mbta6)
```

There is an odd max value for the Boat data. Investigate distributions further with *hist* or *boxplot*.

```{r}
hist(mbta6$Boat)
```

There is a clear outlier - knowing that there are no months in which ridership significantly varied in Boston (i.e. Olympics), this is likely data entry error. Locate the error - infer that it is an added zero and correct it.

```{r}
# Find the row number of the incorrect value: i
i <- which(mbta6$Boat == 40)

# Replace the incorrect value with 4
mbta6$Boat[i] <- 4

# Generate a histogram of Boat column
hist(mbta6$Boat)

```

Investigate seasonal trends in the data.
```{r}
library(ggplot2)

# Look at Boat ridership over time 
ggplot(mbta6, aes(x = month, y = Boat)) +  geom_point() + 
  scale_x_discrete(name = "Month") + 
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")

```

There is a clear seasonal peak to boat ridership.